---
title: "06: Stratifying on time for cohort studies"
author: "Thiago Cerqueira Silva"
format: 
  html: 
    toc: true
    toc-location: right
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
    df-print: kable
execute: 
  warning: false
  message: false
editor: visual
sidebar: home
---

```{r}
# read "dta" files
library(haven) 
# create table
library(gtsummary)
# data wrangling and other functions
library(tidyverse)
# if you want to have a output similar to stata
library(tidylog)
#cox
library(survival)
# plots survival
library(ggsurvfit)
# Limit significant digits to 2, remove scientific notation
options(digits = 2, scipen = 9)
```

# Q1

Open a log file, change to the directory where your ASME data have been copied and then read the Whitehall data, whitehal.dta. In this practical we are interested in the effect of job grade on CHD mortality.

```{r}
whitehall <- read_stata("datasets/WHITEHAL.DTA")

# data wrangling
whitehall <- whitehall |> 
  mutate(
    id = as.integer(id),
    all = as.integer(all), # better keep numeric (not a problem in this case)
    chd = as.integer(chd),
    grade4 = case_when( # recoding as character using case_when, most models will recognize characters as factors with some exemptions (mgcv doesnt)
      grade4 == 1 ~ "admin",
      grade4 == 2 ~ "profess",
      grade4 == 3 ~ "clerical",
      grade4 == 4 ~ "other"
    ),
    smok = case_when(
      smok == 1 ~ "never",
      smok == 2 ~ "ex",
      smok == 3 ~ "1-14/day",
      smok == 4 ~ "15-24/day",
      smok == 5 ~ "25+/day"
    ),
    grade = case_when(
      grade == 1 ~ "higher",
      grade == 2 ~ "lower"),
    cholgrp = as.factor(cholgrp),
    sbpgrp = as.factor(sbpgrp))
```

# Q2

Compute the rates for the two grades of employment categories using the commands below. `pyears()`

**In days**

```{r, results='asis'}
# Calculate rates
pyears(Surv(
  time = as.numeric(timein),
  time2 = as.numeric(timeout),
  event = chd
) ~ grade, data = whitehall, scale = 1) |>
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```

**In years**

```{r, results='asis'}
# Calculate rates
pyears(Surv(
  time = as.numeric(timein),
  time2 = as.numeric(timeout),
  event = chd
) ~ grade, data = whitehall) |>
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```

::: {.callout-tip collapse="true"}
## Coding explanation

Calculate rates stratified by exposure (the two grades of employment: `grade`).

You create a survival object with `Surv()`; it contains duration of follow-up and status at end of follow-up. You dont need to setup it globally as in STATA. You have freedom to use different timescales for each model

You then calculate stratified rates `pyears()`: in the formula, you use a Surv object, and then the stratification; you then pipe this into `summary()`

The argument "scale" from `pyears()` is to scaling the results, for example if your time is in days and want to report in years use scale = 365.25 (the default value)
:::

Calculate the RR by hand (from the two rates shown on the screen).

::: {.callout-note collapse="true"}
## Answer

$\frac{8.8}{4.4}=2$
:::

# Q3

Now, set again the time and outcome variables with stset but this time, use the time of birth as the origin, i.e. organise the data according to current age.

```{r, results='asis'}
# Calculate rates
pyears(Surv(
  time = as.numeric(timein),
  time2 = as.numeric(timeout),
  origin = as.numeric(timebth),
  event = chd
) ~ grade, data = whitehall) |> 
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```

```{r}
# I dont know any easy alternative to see the summary (similar to STATA)
summary(Surv(as.numeric(whitehall$timein) / 365.25,
  as.numeric(whitehall$timeout) / 365.25,
  origin = as.numeric(whitehall$timebth) / 365.25,
  whitehall$chd
))
```

# Q4, Q5 and Q6

Next we need to split the individual follow-up times into intervals specific to different agebands.

Use the stsplit command to create 5-years groups of current age between age 50 and 80 and 10-year groups for the youngest and oldest groups.

```{r, results='asis'}
# dividing by 365.25 to transform in years instead days
# the variables used in time, time2,origin will be removed from the new dataset
white_split <- survSplit(
  Surv(
    time = as.numeric(timein) / 365.25,
    time2 = as.numeric(timeout) / 365.25,
    event = chd,
    origin = as.numeric(timebth) / 365.25
  ) ~ .,
  data = whitehall,
  cut = c(40, seq(50, 80, 5), 90),
  episode = "ageband"
)

white_split |>  filter(id == "5001") |> 
  select(id,
         tstart, tstop, ageband) |> 
  left_join(whitehall |> 
              select(id,timein,timeout,timebth) |> 
              mutate(age_enter = (timein-timebth)/365.25)) |> 
  gt::gt()
```

# Q7

Note that there is no change in the information about length of follow-up

```{r, results='asis'}
# Stratify by grade
# use scale =1 because the time is already in years
pyears(Surv(tstart, tstop,chd) ~ grade,
       data = white_split,
       scale=1) %>%
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```

# Q8

Tabulate by agebands now

```{r}
pyears(Surv(tstart, tstop,chd) ~ ageband,
       data = white_split,
       scale=1) |> 
  summary(n = T, rate = T, ci.r = T, scale = 1000) #use n=T to see the number of individuals who were followed in each ageband
```

# Q9

Calculate the ageband-specific RRs for Low grade versus High grade using stmh and assess whether the effect of grade is the same across all the strata. Is it necessary to report the stratum specific estimates? Do you think that age confounds the relationship between job grade and CHD mortality?

```{r}
pyears(Surv(tstart, tstop,chd) ~ ageband+grade,
       data = white_split,
       scale=1)|> 
  summary(n = T, rate = T, ci.r = T, scale = 1000) 

# To get the RR of grade adjusted for ageband, we need additional steps

dt_rates <- pyears(Surv(tstart, tstop,chd) ~ ageband+grade,
       data = white_split,
       scale=1,
       data.frame = T) 
# ageband is coded as numeric in the output
dt_rates$data$ageband <- as.factor(dt_rates$data$ageband)
dt_mh <- dt_rates$data |> arrange(desc(grade))
```

Painful to do MH in R

```{r}
dat.tab04 <- sapply(2:length(unique(dt_mh$ageband)), function(x) 
   as.matrix(dt_mh[dt_mh$ageband == x,c(5,3)], ncol = 2, byrow = TRUE), 
   simplify = "array")
epiR::epi.2by2(dat = dat.tab04, method = "cohort.time", digits = 2, 
   conf.level = 0.95, interpret = FALSE, 
   outcome = "as.columns")

```

Model based is way better

```{r}
mod1 <-  glm(event ~ grade + ageband + offset(log(pyears)),
      family = poisson,
      data= dt_rates$data) 
mod_int <-  glm(event ~ grade*ageband + offset(log(pyears)),
      family = poisson,
      data= dt_rates$data) 
tbl_regression(mod1,exponentiate = T)

```

::: {.callout-tip collapse="true"}
## Explanation - Offset

The Poisson regression usually is used for model count data, but we can also model rates with it. To do that, we need to transform the count in rates, and the offset variable will inform us about the "size" of exposure of each individual/unit. The regression coefficient for an offset variable is constrained to be 1, thus allowing our model to represent rates rather than counts. When you fit a model with an offset, the exponentiated regression coefficient of a predictor variable tells us how much the expected rate changes multiplicatively for a one unit increase in the predictor variable. When using an offset, the assumption is made that doubling the unit size (measurement time, etc.) will lead to a doubling of the count outcome. If this assumption is not appropriate, controlling for the unit size as a covariate instead of an offset may be more appropriate. A likelihood ratio test could be used to determine if the fit of these two models is significantly different.

Why the offset is in log? Consider the rate $\lambda$

$$\lambda=\frac{\mu_{i}}{t_{i}}$$

$$ log(\lambda)=\alpha +\beta x $$

The model will be $$ log(\mu)-log(t)=\alpha +\beta x $$

$$ log(\mu)=\alpha +\beta x +log(t) $$
:::

Check interaction

```{r}
lmtest::lrtest(mod1,mod_int)
```

Different answers from STATA output. Here, we are modelling it using Poisson and testing interaction using a likelihood test. Instead, in the Stata solutions it is with stratified rate ratios (Mantel-Haenszel)

All subsequent questions will be answered using model based / likelihood ratio tests for interactions

# Q10

Using strate and stmh examine (a) the effect of smoking on CHD mortality, (b) whether smoking confounds the relationship between job grade and CHD mortality. You may wish to recode the smoking variable into three categories (i) never, (ii) ex, (iii) current smokers

```{r}

# Recode
white_split <- white_split |> 
  mutate(smok3 = as.factor(case_when(smok == "never" ~ "never",
                                     smok == "ex" ~ "ex",
                                     smok == "1-14/day" ~ "current",
                                     smok == "15-24/day" ~ "current",
                                     smok == "25+/day" ~ "current")),
         smok3 = fct_relevel(smok3, "never", "ex", "current"))# Order level



pyears(Surv(tstart, tstop,chd) ~ smok3+grade,
       data = white_split,
       scale=1)|> 
  summary(n = T, rate = T, ci.r = T, scale = 1000) 

# To get the RR of grade adjusted for ageband, we need additional steps

dt_rates <- pyears(Surv(tstart, tstop,chd) ~ smok3+grade,
       data = white_split,
       scale=1,
       data.frame = T) 
dt_mh <- dt_rates$data 
```

```{r}
mod1 <-  glm(event ~ grade + smok3 + offset(log(pyears)),
      family = poisson,
      data= dt_rates$data) 
mod_int <-  glm(event ~ grade*smok3 + offset(log(pyears)),
      family = poisson,
      data= dt_rates$data) 
tbl_regression(mod1,exponentiate = T)

```

Check interaction

```{r}
lmtest::lrtest(mod1,mod_int)
```

# Q11

Examine the effect of job grade on CHD mortality adjusted for ageband and smoking simultaneously. What do you conclude about the effect of job grade on CHD mortality having adjusted for both of these factors together?

```{r}
pyears(Surv(tstart, tstop,chd) ~ smok3+grade+ageband,
       data = white_split,
       scale=1)|> 
  summary(n = T, rate = T, ci.r = T, scale = 1000) 

dt_rates <- pyears(Surv(tstart, tstop,chd) ~ smok3+grade+ageband,
       data = white_split,
       scale=1,
       data.frame = T)
# ageband is coded as numeric in the output
dt_rates$data$ageband <- as.factor(dt_rates$data$ageband)
mod1 <-  glm(event ~ grade+smok3 + ageband+offset(log(pyears)),
      family = poisson,
      data= dt_rates$data) 
tbl_regression(mod1,exponentiate = T)
```

```{r}
mod_int <-  glm(event ~ grade*smok3 + ageband+offset(log(pyears)),
      family = poisson,
      data= dt_rates$data) 
lmtest::lrtest(mod1,mod_int)
```

::: {.callout-note collapse="true"}
## Answer

There is no evidence of effect modification between grade and smoking. $\chi_{2}^2 = 1.25$ and p=0.53
:::

# Q12

The file whchd.dta holds the Whitehall data already expanded by current age and period (in 5-years periods) and has been stset for the analysis of CHD events. It also holds a variable, called rate, giving the corresponding age and period specific CHD mortality rates for England and Wales (per 1,000,000 person years). Compute the CHD SMR for the cohort of Whitehall civil servants

```{r}
# way more complicated than STATA (using this format of data)
# the stranges names from stata dont work verwy well in R - use clean_names to fix
whcd <- haven::read_dta("datasets/WHCHD.DTA") |> janitor::clean_names()
# get the expected number of cases - adjusted by period, ageband and grade

expectdt <- whcd |> group_by(ageband, grade, period) |> 
  reframe(a=mean(rate)/1e6,
          totalpy=sum(t-t0)) |> 
  reframe(e=a*totalpy) |> 
  reframe(sum(e))


observedt <- whcd |> 
  reframe(sum(chd, na.rm=T))

popEpi::poisson.ci(147,227)
```
